---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# czso

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/czso)](https://CRAN.R-project.org/package=czso)
[![Travis build status](https://travis-ci.org/petrbouchal/czso.svg?branch=master)](https://travis-ci.org/petrbouchal/czso)
<!-- badges: end -->

The goal of czso is to provide direct, programmatic, hassle-free access from R to open data provided by the Czech Statistical Office (CZSO).

This is done by

1. **providing direct access from R to the catalogue of open CZSO datasets**, eliminating the hassle from data discovery. Normally this is done done through [the CZSO's product catalogue](https://www.czso.cz/csu/czso/otevrena-data-v-katalogu-produktu-csu) which is unfortunately a bit clunky, or [data.gov.cz](https://data.gov.cz), which is not a natural starting point for many.

2. **providing a function to load a specific dataset to R** directly from the CZSO's datastore, eliminating the friction of copying a URL, downloading, unzipping etc.

Additionally, the package provides metadata on datasets.

## Installation

You can install the latest release from [github](https://github.com/petrbouchal/czso) with:

``` r
remotes::install_github("petrbouchal/czso", ref = github_release())
```

or the latest version with:

``` r
remotes::install_github("petrbouchal/czso")
```

## Example

Imagine you are looking for a dataset whose title refers to wages (mzda/mzdy):

```{r example}
library(czso)

# first, retrieve the list of available CZSO datasets, filtering for mzda/mzdy
get_czso_catalogue(title_filter = "mzd[ay]")
```

We can see the `czso_id` for the required dataset - now use it to get the dataset:

```{r example-cont}
get_czso_table("110080")
```

Alternatively, you could store the whole CZSO catalogue in an object and filter yourself. This is especially useful if you expect to need multiple tries.

```{r example-alt}
library(dplyr, warn.conflicts = F)
library(stringr, warn.conflicts = F)
catalogue <- get_czso_catalogue()

catalogue %>% 
  filter(str_detect(title, "mzda"))
```

The latter allows you to search through the list - or simply look through it - without the overhead of reusing the `get_dataset()` function which downloads and transforms the underlying data.

## Credit and notes

- not created or endorsed by the Czech Statistical Office, though they, as well as [the open data team at the Ministry of Interior](https://data.gov.cz/) deserve credit for getting the data out there.
- the package relies on the data.gov.cz catalogue of open data and on the CZSO's local catalogue
- NB: The robots.txt at the domain hosting the CZSO's catalogue prohibits robots from accessing it; while this may be an inappropriate/erroneous setting for what is in essence a data API, this package tries to honor the spirit of that setting by only accessing the API once per `get_table()` call, relying on a different system for `get_catalogue()`. Hence, *do not use this package for harvesting large numbers of datasets from the CZSO.*

## See also

This package takes inspiration from the packages

- [eurostat](https://github.com/rOpenGov/eurostat/)
- [OECD](https://github.com/expersso/OECD)

which are very useful in their own right - much recommended.

